---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# About
This vignette shows the basic usage of the **vasp2R** package. First we will import an example CHG file, do some manipulation and finally plot the result of the VASP calculation. 

The VASP files used to create the CHG file supplied by this package are taken from the [third Hands-on example of VASP's documentation](https://www.vasp.at/vasp-workshop/slides/Handson3.tgz). I used the first one: the clean Ni surface. Why did I used the CHG and not the CHGCAR? Because it also contains the required information but is smaller in size and I want to keep package's size at a minimum.

# Import
First of all you have to be sure of have the **vasp2R** package installed.
```{r, cache = TRUE }
check.validity <- try( library( "vasp2R" ), silent = TRUE )
if ( class( check.validity ) == "try-error" ){
    ## the package is not installed yet
    devtools::install_github( "theGreatWhiteShark/vasp2R" )
    library( vasp2R )
}
```

During installation the CHG files is copied as one of the package's assets on the hard disk. We will import it into R using the *vasp.import* function.

```{r, cache = TRUE}

imported.data <- vasp.import( system.file( "example/CHG",
                                          package = "vasp2R" ) )
```

# Manipulation and visualization

Afterwards we will reproduce the imported VASP results of a single Ni atom with periodic boundary conditions to make it actually look like a surface


Then we establish bonds between the individual atoms (for a better visualization of the results).


As a final step we can rotate the positions to change our point of view. For such a clean and symmetric surface this is a little bit useless. But for more complex structures it will become quite handy.
